UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-05-27 ms/main.tex
clarify some stuff and better image captions

\newcommand{\beginsupplement}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }
\newcommand{\sgcomment}[1]{{\textcolor{blue}{SG: #1}}}
\newcommand{\adpcomment}[1]{{\textcolor{orange}{ADP: #1}}}
%\usepackage{subfig}
\usepackage{adjustbox}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{float}
\usepackage{dblfloatfix}
\usepackage[switch]{lineno}
\usepackage{comment}
%\usepackage[section]{placeins}

@@

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=0.7\textwidth]{images/montage_1000G_labels_axes.pdf}
    \caption{Four methods of dimension reduction of 1KGP genotype data with population labels (i) PCA maps individuals in a triangle with vertices corresponding to African, Asian/Native American, and European continental ancestry. Discarding lower-variance PCs leads to overlap of populations with no close affinity, such as Central and South American populations with South Asians. (ii) t-SNE forms groups corresponding to continents, with some overlap between European and Central and South American people. Smaller subgroups are visible within continental clusters. The cloud of peripheral points results from the method's poor convergence. (iii) UMAP forms distinct clusters related to continent with clearly defined subgroups. Japanese, Finnish, Luhya, and some Punjabi and Telugu populations form separate clusters consistent with their population history\cite{10002015global}. (iv) UMAP on the first 15 principal components forms fine-scale clusters for individual populations. Groups closely related by ancestry or geography, such as African Caribbean/African American, Spanish/Italian, and Kinh/Dai populations cluster together. Results using t-SNE on principal components are presented in figure~\ref{fig:supp_megamontage_pc2_9}. Axes in UMAP and t-SNE are arbitrary. Since the algorithms prioritize local distances, long distances between clusters are not meaningful.
    ACB, African Caribbean in Barbados;
    ASW, African Ancestry in Southwest US;
    BEB, Bengali;
    CDX, Chinese Dai;
    CEU, Utah residents with Northern/Western European ancestry;
    CHB, Han Chinese;
    CHS, Southern Han Chinese;
    CLM, Colombian in Medellin, Colombia;
    ESN, Esan in Nigeria;
    FIN, Finnish;
    GBR, British in England and Scotland;
    GWD, Gambian;
    GTH, Gujarati;
    IBS, Iberian in Spain;
    ITU, Indian Telugu in the UK;
    JPT, Japanese;
    KHV, Kinh in Vietnam;
    LWK, Luhya in Kenya;
    MSL, Mende in Sierra Leone;
    MXL, Mexican in Los Angeles, California;
    PEL, Peruvian;
    PJL, Punjabi in Lahore, Pakistan;
    PUR, Puerto Rican;
    STU, Sri Lankan Tamil in the UK;
    TSI, Tuscani in Italy;
    YRI, Yoruba in Nigeria}
    \label{fig:1000g_compare}
\end{figure*}

@@

\begin{figure*}
    \centering
    \begin{subfigure}{0.5\columnwidth}
    \includegraphics[width=\columnwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.pdf}
    \caption{Subset of Hispanic population of the HRS}
    \label{fig:supp_umap_hrs_hisp_1kgp_hisp}
    \end{subfigure}\hfill
        \begin{subfigure}{0.5\columnwidth}
    \includegraphics[width=\columnwidth]{images/Asthma_FullAsian_PC_QC_new_UMAP_PC8_NC2_NN15_MD05_20193117303txt_eth_for_ms_lw2.jpeg}
    \caption{Subset of the Asian population of the UKBB}
        \label{fig:umap_ukbb_indian}
    \end{subfigure}
    \caption{Applying UMAP to subets of data can reveal deep population structure. (a) UMAP on the top $7$ principal components of the self-identified Hispanic population of the HRS reveals a cluster (highlighted). Colouring the points by birthplace shows they were born almost entirely in the Mountain region of the United States (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana). When populations from the 1KGP are projected onto the UMAP embedding they do not map to the cluster. Six 1KGP populations are presented:
    CLM, Colombian in Medellin, Colombia;
    IBS, Iberian in Spain;
    MXL, Mexican in Los Angeles, California;
    PEL, Peruvian;
    PUR, Puerto Rican;
    TSI, Tuscani in Italy. Figure~\ref{fig:supp_umap_hrs_hisp_admix} presents the same projection of individuals from the HRS coloured by estimated admixture proportions. (b) UMAP on the top $8$ principal components of the self-identified Asian populations of the UKBB create clusters. Almost all individuals in the highlighted cluster are Indian individuals born in Kenya.}
    \label{fig:subpops}
\end{figure*}

@@

\begin{figure*}%
\centering
\begin{subfigure}{.5\columnwidth}
\includegraphics[width=\columnwidth]{images/UKBB_PC0_PC1_eth_crop.jpeg}%
\caption{Principal components 1 and 2}%
\label{fig:pc_ukbb}%
\end{subfigure}\hfill%
\begin{subfigure}{.5\columnwidth}
\includegraphics[width=\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_eth_labels.pdf}%
\caption{UMAP on first 10 principal components}%
\label{fig:umap_ukbb_pc10}%
\end{subfigure}%
\caption{The UKBB coloured by self-reported ethnic background. (a) The first two principal components, showing the usual triangle with vertices corresponding to African, Asian/Native American, and European ancestries, and intermediate values indicating admixture or lack of relationship to the vertex populations. (b) UMAP on the first 10 principal components. The cluster of White British and White Irish individuals is greatly expanded, with the Irish forming a distinct sub cluster mixed with the White British population. South Asian and East Asian individuals form their separate clusters, as do individuals of African or Caribbean backgrounds. Population clusters are connected by "trails" comprised of large proportions of individuals with mixed backgrounds.
BA, Black African; 
BC, Black Caribbean; 
BG, Bangladeshi; 
CHN, Chinese; 
IND, Indian;
PK, Pakistani;
WB, White British;
WI, White Irish;
WBC, White and Black Caribbean; 
WBA, White and Black African; 
WAA, White and Asian;
AAB, Any other Asian Background; 
ABB, Any other Black Background;
AWB, Any other White Background;
AMB, Any other Mixed Background;
OEG, Other ethnic group. 
}
\label{fig:fig_ukbb}
\end{figure*}

@@

\begin{figure}
    \centering
    \begin{subfigure}{\columnwidth}
    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ns_permuted_10nn_sd50000_base50000_201883118582.pdf}
    \includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ew_permuted_10nn_sd50000_base50000_201883118582.pdf}
    \caption{Colouring the projection by geographic coordinates.}
    \label{fig:umap_ukbb_geo_osgb}
    \end{subfigure}
     \begin{subfigure}{\columnwidth}
         \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714161841_Height_res_pct1_f.pdf}
    \includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416519_leukocyte_count_pct5_f.pdf}
   \caption{Colouring the projection by phenotypic data}
    \label{fig:umap_height_female}
      \end{subfigure}
      \caption{Relationships between population structure, geography, and phenotype heterogeneity. (Top) Each individual is coloured by their geographical coordinates of residence. Coordinates follow the UKBB's OSGB1936 geographic grid system and represent distance from the Isles of Scilly, which lie southwest of Great Britain. The left image colours individuals by their north-south ("northing") coordinates, and the right image colours them by their east-west ("easting") coordinates. Adding more components creates finer clusters.  (figures~\ref{fig:supp_montage_ukbb_ns} and~\ref{fig:supp_montage_ukbb_ew}). Northing values were truncated between $100$km and $700$km, and easting values were truncated between $200$km and $600$km. (Bottom) Females from the same projection coloured by age-adjusted difference from mean population height (left) and leukocyte counts (right). Individuals with missing data were excluded. To protect participant privacy, data in these images has been randomized as explained in the materials and methods section.}
\end{figure}

@@

\begin{figure}
    \centering
    \begin{subfigure}{\columnwidth}
    \includegraphics[width=0.5\columnwidth]{images/UKBB_mapsInMaps_2D3_permute50.jpg}
    \caption{Each point is an individual placed based on where they live. Patterns in genetic similarity are visible in Scotland, South England, North and South Wales, the East and West Midlands, and major urban centres.}
    \label{fig:umap_ukbb_geo_map}
    \end{subfigure}
     \begin{subfigure}{\columnwidth}
         \includegraphics[width=0.9\columnwidth]{images/meanWorldMap.jpg}
   \caption{Geographic distribution of UMAP coordinates. Using the country of birth of individuals in the UKBB, we colour countries by the closeness in 3D UMAP space of those born there. Broad patterns of similarity appear in East Asia, South Asia, North African and the Middle East, West Africa, and South America. Differences between neighbouring countries can reflect both ancient population structure and recent differences in migration history. Evidence of migrations related to colonialism are visible with, e.g., European ancestry in South Africa and South Asian ancestry in Kenya and Tanzania. Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States \sgcomment{rationale?}. Zoomed maps of East Asia, the Caribbean, and Europe are available in figures~\ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur}, respectively.}
    \label{fig:umap_world_map}
      \end{subfigure}
      \caption{Maps coloured by 3D UMAP projections of the top 20 principal components of the UKBB. Each individual is assigned a 3D RGB vector based on 3D UMAP coordinates. Individuals who are closer to each other in the projection will be closer in colour in the maps. More details on colouring, as well as randomization of points to protect participant privacy, are available in the materials and methods. \sgcomment{Merge captions. Make sure to have overall title! }}
\end{figure}

@@

Using UMAP reveals intriguing clusters that would otherwise have been very difficult to identify via numerous pairwise PCA plots, such as the geographically restricted cluster within the Hispanic population of the HRS, or the splits within the Gujarati and Punjabi population samples in the 1kGP that were not visible through PCA or Admixture analysis alone \sgcomment{How did the Reich paper find it?}. Application to the UKBB underscores the strength of UMAP in large cohorts, with fine relationships between genotype and phenotype and geography presented in visualizations that account for natural genetic clustering. Figures~\ref{fig:umap_height_female}, \ref{fig:supp_ukbb_fev_f}, and \ref{fig:supp_ukbb_zoom} demonstrate phenotypic variation within and across clusters, with height showing continuous variation across admixture edges, as expected from genetically controlled traits, and others, such as leukocyte counts or FEV1, showing sharper boundaries, as expected from environmentally determined traits.

UMAP works well at highlighting groups of individuals who share more ancestry; when clusters form consistently, this suggests the individuals within it share more ancestry, though nothing can be said about the differences between clusters. The 1KGP dataset was deliberately crafted to come from geographically separated populations that are homogeneous internally. Thus, most of its populations form isolated clusters, with the exception of those that most closely share ancestries. \sgcomment{I don't understand what this paragraph is trying to say. You might be a bit too wordy here. }

@@

Colours for maps in figures~\ref{fig:umap_world_map}, \ref{fig:umap_ukbb_geo_map}, \ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur} were determined by projecting data to 3D and using each 3D coordinate as an RGB coordinate. For the world map, countries were determined using the country of birth variable, with a country's colour being determined by the mean $x$, $y$, and $z$ values of all individuals born in that country. Because many self-identified White British individuals were born abroad, including them everywhere would skew the colour scheme; they were included only if they were born in the UK, Europe, Australia, Canada, or the United States. This approach to colouring is sensitive to sample sizes as UMAP will give more space to larger populations. For this reason, differences between Ireland/the UK and other countries appear exaggerated. Selecting a smaller minimum distance can result in breaking connections between clusters, leading to different colour groups appearing. We provide an example world map using 40 principal components and a minimum distance of 0.001 in figure~\ref{fig:supp_umap_ukbb_worldmap_alt}. \sgcomment{Is this really a "material and method" comment? THe logical flow with the last comment is not clear to me.}

@@

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/meanAsiaMap.jpg}
    \end{subfigure}
    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on Asia. \sgcomment{add caption from fig5b? No cost to being self-contained here. }}
    \label{fig:supp_umap_ukbb_asia}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/meanCarribeanMap.jpg}
    \end{subfigure}
    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on the Caribbean. \sgcomment{add caption from fig5b? No cost to being self-contained here. }}
    \label{fig:supp_umap_ukbb_car}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/meanEuroMap.jpg}
    \end{subfigure}
    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on Europe. \sgcomment{add caption from fig5b? No cost to being self-contained here. }}
    \label{fig:supp_umap_ukbb_eur}
\end{figure*}

>>>>

\newcommand{\beginsupplement}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }
\newcommand{\sgcomment}[1]{{\textcolor{blue}{SG: #1}}}
\newcommand{\adpcomment}[1]{{\textcolor{orange}{ADP: #1}}}

\newcommand{\rulesep}{\unskip\ \vrule\ }

%\usepackage{subfig}
\usepackage{adjustbox}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{float}
\usepackage{dblfloatfix}
\usepackage[switch]{lineno}
\usepackage{comment}
%\usepackage[section]{placeins}

@@

\begin{figure*}[!ht]
    \centering
    \includegraphics[width=0.7\textwidth]{images/montage_1000G_labels_axes.pdf}
    \caption{\textbf{Four methods of dimension reduction of 1KGP genotype data with population labels.} (i) PCA maps individuals in a triangle with vertices corresponding to African, Asian/Native American, and European continental ancestry. Discarding lower-variance PCs leads to overlap of populations with no close affinity, such as Central and South American populations with South Asians. (ii) t-SNE forms groups corresponding to continents, with some overlap between European and Central and South American people. Smaller subgroups are visible within continental clusters. The cloud of peripheral points results from the method's poor convergence. (iii) UMAP forms distinct clusters related to continent with clearly defined subgroups. Japanese, Finnish, Luhya, and some Punjabi and Telugu populations form separate clusters consistent with their population history\cite{10002015global}. (iv) UMAP on the first 15 principal components forms fine-scale clusters for individual populations. Groups closely related by ancestry or geography, such as African Caribbean/African American, Spanish/Italian, and Kinh/Dai populations cluster together. Results using t-SNE on principal components are presented in figure~\ref{fig:supp_megamontage_pc2_9}. Axes in UMAP and t-SNE are arbitrary. Since the algorithms prioritize local distances, long distances between clusters are not meaningful.
    ACB, African Caribbean in Barbados;
    ASW, African Ancestry in Southwest US;
    BEB, Bengali;
    CDX, Chinese Dai;
    CEU, Utah residents with Northern/Western European ancestry;
    CHB, Han Chinese;
    CHS, Southern Han Chinese;
    CLM, Colombian in Medellin, Colombia;
    ESN, Esan in Nigeria;
    FIN, Finnish;
    GBR, British in England and Scotland;
    GWD, Gambian;
    GTH, Gujarati;
    IBS, Iberian in Spain;
    ITU, Indian Telugu in the UK;
    JPT, Japanese;
    KHV, Kinh in Vietnam;
    LWK, Luhya in Kenya;
    MSL, Mende in Sierra Leone;
    MXL, Mexican in Los Angeles, California;
    PEL, Peruvian;
    PJL, Punjabi in Lahore, Pakistan;
    PUR, Puerto Rican;
    STU, Sri Lankan Tamil in the UK;
    TSI, Tuscani in Italy;
    YRI, Yoruba in Nigeria}
    \label{fig:1000g_compare}
\end{figure*}

@@

\begin{figure*}
    \centering
    \begin{subfigure}{0.5\columnwidth}
    \includegraphics[width=\columnwidth]{images/HRS_1000G_NP1_UMAP_PC7_NC2_NN15_MD05_pca_hrshisp_added1kgp_2018115153245_1kgp_hisp.pdf}
    \caption{}
    \label{fig:supp_umap_hrs_hisp_1kgp_hisp}
    \end{subfigure}\hfill
        \begin{subfigure}{0.5\columnwidth}
    \includegraphics[width=\columnwidth]{images/Asthma_FullAsian_PC_QC_new_UMAP_PC8_NC2_NN15_MD05_20193117303txt_eth_for_ms_lw2.jpeg}
    \caption{}
        \label{fig:umap_ukbb_indian}
    \end{subfigure}
    \caption{\textbf{Applying UMAP to subets of data can reveal deep population structure.} \textbf{(a)} UMAP on the top $7$ principal components of the self-identified Hispanic population of the HRS reveals a cluster (highlighted). Colouring the points by birthplace shows they were born almost entirely in the Mountain region of the United States (New Mexico, Arizona, Colorado, Utah, Nevada, Wyoming, Idaho, and Montana). When populations from the 1KGP are projected onto the UMAP embedding they do not map to the cluster. Six 1KGP populations are presented:
    CLM, Colombian in Medellin, Colombia;
    IBS, Iberian in Spain;
    MXL, Mexican in Los Angeles, California;
    PEL, Peruvian;
    PUR, Puerto Rican;
    TSI, Tuscani in Italy. Figure~\ref{fig:supp_umap_hrs_hisp_admix} presents the same projection of individuals from the HRS coloured by estimated admixture proportions. \textbf{(b)} UMAP on the top $8$ principal components of the self-identified Asian populations of the UKBB create clusters. Almost all individuals in the highlighted cluster are Indian individuals born in Kenya.}
    \label{fig:subpops}
\end{figure*}

@@

\begin{figure*}%
\centering
\begin{subfigure}{.5\columnwidth}
\includegraphics[width=\columnwidth]{images/UKBB_PC0_PC1_eth_crop.jpeg}%
\caption{}%
\label{fig:pc_ukbb}%
\end{subfigure}\hfill%
\begin{subfigure}{.5\columnwidth}
\includegraphics[width=\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_eth_labels.pdf}%
\caption{}%
\label{fig:umap_ukbb_pc10}%
\end{subfigure}%
\caption{\textbf{The UKBB coloured by self-reported ethnic background.} \textbf{(a)} The first two principal components, showing the usual triangle with vertices corresponding to African, Asian/Native American, and European ancestries, and intermediate values indicating admixture or lack of relationship to the vertex populations. \textbf{(b)} UMAP on the first 10 principal components. The cluster of White British and White Irish individuals is greatly expanded, with the Irish forming a distinct sub cluster mixed with the White British population. South Asian and East Asian individuals form their separate clusters, as do individuals of African or Caribbean backgrounds. Population clusters are connected by "trails" comprised of large proportions of individuals with mixed backgrounds.
BA, Black African; 
BC, Black Caribbean; 
BG, Bangladeshi; 
CHN, Chinese; 
IND, Indian;
PK, Pakistani;
WB, White British;
WI, White Irish;
WBC, White and Black Caribbean; 
WBA, White and Black African; 
WAA, White and Asian;
AAB, Any other Asian Background; 
ABB, Any other Black Background;
AWB, Any other White Background;
AMB, Any other Mixed Background;
OEG, Other ethnic group. 
}
\label{fig:fig_ukbb}
\end{figure*}

@@

\begin{figure}
    \centering
    \begin{subfigure}{\columnwidth}
    \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ns_permuted_10nn_sd50000_base50000_201883118582.pdf}
    \includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_ew_permuted_10nn_sd50000_base50000_201883118582.pdf}
    \caption{}
    \label{fig:umap_ukbb_geo_osgb}
    \end{subfigure}
     \begin{subfigure}{\columnwidth}
         \includegraphics[width=0.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_2018714161841_Height_res_pct1_f.pdf}
    \includegraphics[width=.4\columnwidth]{images/UKBB_UMAP_PC10_NN15_MD05_2018328174511_201871416519_leukocyte_count_pct5_f.pdf}
   \caption{}
    \label{fig:umap_height_female}
      \end{subfigure}
      \caption{\textbf{UMAP captures relationships between population structure, geography, and phenotype heterogeneity.} \textbf{(a)} Each individual is coloured by their geographical coordinates of residence. Coordinates follow the UKBB's OSGB1936 geographic grid system and represent distance from the Isles of Scilly, which lie southwest of Great Britain. The left image colours individuals by their north-south ("northing") coordinates, and the right image colours them by their east-west ("easting") coordinates. Adding more components creates finer clusters.  (figures~\ref{fig:supp_montage_ukbb_ns} and~\ref{fig:supp_montage_ukbb_ew}). Northing values were truncated between $100$km and $700$km, and easting values were truncated between $200$km and $600$km. \textbf{(b)} Females from the same projection coloured by age-adjusted difference from mean population height (left) and leukocyte counts (right). Individuals with missing data were excluded. To protect participant privacy, data in these images has been randomized as explained in the materials and methods section.}
\end{figure}

@@

%\begin{figure}[h!]
%  \centering
%  \begin{subfigure}[b]{0.4\linewidth}
%    \includegraphics[width=\linewidth]{Figures/1KGP_UMAP_labels.jpeg}
%  \end{subfigure}
%  \rulesep
%  \begin{subfigure}[b]{0.4\linewidth}
%    \includegraphics[width=\linewidth]{Figures/1KGP_PC15_UMAP_labels.jpeg}
%  \end{subfigure}
%  \caption{UMAP embeddings of the 1KGP genotype data (left) and the top 15 principal components of the genotype data (right). Each point is coloured and labelled by its population. See Appendix~\ref{AppendixA} for label definitions.}
%  \label{fig:1KGP}
%\end{figure}

%\begin{figure}
%    \centering
%     \textbf{Test}\par\medskip
%    \begin{subfigure}{\columnwidth}
%    \includegraphics[width=0.35\columnwidth]{images/UKBB_mapsInMaps_2D3_permute50.jpg}
%     %\includegraphics[width=0.55\columnwidth]{images/meanWorldMap.jpg}
%    \caption{Each point is an individual placed based on where they live. Patterns in genetic similarity are visible in Scotland, South England, North and South Wales, the East and West Midlands, and major urban centres.}
%    \label{fig:umap_ukbb_geo_map}
%    \end{subfigure}
%          \rulesep
%     \begin{subfigure}{\columnwidth}
%         \includegraphics[width=0.45\columnwidth]{images/meanWorldMap.jpg}
%   \caption{Geographic distribution of UMAP coordinates. Using the country of birth of individuals in the UKBB, we colour countries by the closeness in 3D UMAP space of those born there. Broad patterns of similarity appear in East Asia, South Asia, North African and the Middle East, West Africa, and South America. Differences between neighbouring countries can reflect both ancient population structure and recent differences in migration history. Evidence of migrations related to colonialism are visible with, e.g., European ancestry in South Africa and South Asian ancestry in Kenya and Tanzania. Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States. \sgcomment{rationale?}. Zoomed maps of East Asia, the Caribbean, and Europe are available in figures~\ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur}, respectively.}
%    \label{fig:umap_world_map}
%      \end{subfigure}
%      \caption{Maps coloured by 3D UMAP projections of the top 20 principal components of the UKBB. Each individual is assigned a 3D RGB vector based on 3D UMAP coordinates. Individuals who are closer to each other in the projection will be closer in colour in the maps. More details on colouring, as well as randomization of points to protect participant privacy, are available in the materials and methods. \sgcomment{Merge captions. Make sure to have overall title! }}
%\end{figure}

\begin{figure}[h!]
  \centering
  \begin{subfigure}[b]{0.4\linewidth}
    \includegraphics[width=\linewidth]{images/UKBB_mapsInMaps_2D3_permute50.jpg}
    \caption{}
    \label{fig:umap_ukbb_geo_map}
  \end{subfigure}
  \begin{subfigure}[b]{0.4\linewidth}
    \includegraphics[width=\linewidth]{images/meanWorldMap.jpg}
    \caption{}
  \label{fig:umap_world_map}
  \end{subfigure}
  \caption{\textbf{Maps coloured by 3D UMAP projections of the top 20 principal components of the UKBB.} Each individual is assigned a 3D RGB vector based on 3D UMAP coordinates (a flattened projection is in the top right of \textbf{(a)}. Individuals who are closer to each other in the projection will be closer in colour in the maps. More details on colouring, as well as randomization of points to protect participant privacy, are available in the materials and methods. \textbf{(a)} Each point is an individual placed based on where they live. Patterns in genetic similarity are visible in Scotland, South England, North and South Wales, the East and West Midlands, and major urban centres. \textbf{(b)} Geographic distribution of UMAP coordinates. Using the country of birth of individuals in the UKBB, we colour countries by the closeness in 3D UMAP space of those born there. Broad patterns of similarity appear in East Asia, South Asia, North African and the Middle East, West Africa, and South America. Differences between neighbouring countries can reflect both ancient population structure and recent differences in migration history. Evidence of migrations related to colonialism are visible with, e.g., European ancestry in South Africa and South Asian ancestry in Kenya and Tanzania. Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States, where UKBB participants already tended to have European ancestry. \sgcomment{rationale?}. Zoomed maps of East Asia, the Caribbean, and Europe are available in figures~\ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur}, respectively.}
\end{figure}

@@

Using UMAP reveals  clusters that would have been difficult to identify via numerous pairwise PCA plots, such as the geographically restricted cluster within the Hispanic population of the HRS, or the splits within the Gujarati and Punjabi population samples in the 1kGP that were not visible through PCA or Admixture analysis alone \sgcomment{How did the Reich paper find it?}\adpcomment{PCA on the Gujarati alone followed by FST. I guess the highlight is it was found without having to look at a sub-population specifically}. Application to the UKBB underscores the strength of UMAP in large cohorts, with fine relationships between genotype and phenotype and geography presented in visualizations that account for natural genetic clustering. Figures~\ref{fig:umap_height_female}, \ref{fig:supp_ukbb_fev_f}, and \ref{fig:supp_ukbb_zoom} demonstrate phenotypic variation within and across clusters, with height showing continuous variation across admixture edges, as expected from genetically controlled traits, and others, such as leukocyte counts or FEV1, showing sharper boundaries, as expected from environmentally determined traits.

\adpcomment{(scrapped a paragraph here)}

@@

Colours for maps in figures~\ref{fig:umap_world_map}, \ref{fig:umap_ukbb_geo_map}, \ref{fig:supp_umap_ukbb_asia}, \ref{fig:supp_umap_ukbb_car}, and \ref{fig:supp_umap_ukbb_eur} were determined by projecting data to 3D and using each 3D coordinate as an RGB coordinate. For the world map, countries were determined using the country of birth variable, with a country's colour being determined by the mean $x$, $y$, and $z$ values of all individuals born in that country. Because many self-identified White British individuals were born abroad, including them everywhere would skew the colour scheme; they were included only if they were born in the UK, Europe, Australia, Canada, or the United States. This approach to colouring is sensitive to sample sizes as UMAP will give more space to larger populations. For this reason, differences between Ireland/the UK and other countries appear exaggerated. Selecting a smaller minimum distance can result in breaking connections between clusters, leading to different colour groups appearing. We provide an example world map using 40 principal components and a minimum distance of 0.001 in figure~\ref{fig:supp_umap_ukbb_worldmap_alt}. \sgcomment{Is this really a "material and method" comment? THe logical flow with the last comment is not clear to me.} \adpcomment{This was moved to trim some text from the main body, I can just remove it as it's redundant with info from 2 paragraphs earlier.}

@@

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/meanAsiaMap.jpg}
    \end{subfigure}
    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on Asia. Geographic distribution of UMAP coordinates. Using the country of birth of individuals in the UKBB, we colour countries by the closeness in 3D UMAP space of those born there. Broad patterns of similarity appear in East Asia, South Asia, North African and the Middle East, West Africa, and South America. Differences between neighbouring countries can reflect both ancient population structure and recent differences in migration history. Evidence of migrations related to colonialism are visible with, e.g., European ancestry in South Africa and South Asian ancestry in Kenya and Tanzania. Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States, where UKBB participants already tended to have European ancestry.}
    \label{fig:supp_umap_ukbb_asia}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/meanCarribeanMap.jpg}
    \end{subfigure}
    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on the Caribbean. Geographic distribution of UMAP coordinates. Using the country of birth of individuals in the UKBB, we colour countries by the closeness in 3D UMAP space of those born there. Broad patterns of similarity appear in East Asia, South Asia, North African and the Middle East, West Africa, and South America. Differences between neighbouring countries can reflect both ancient population structure and recent differences in migration history. Evidence of migrations related to colonialism are visible with, e.g., European ancestry in South Africa and South Asian ancestry in Kenya and Tanzania. Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States, where UKBB participants already tended to have European ancestry.}
    \label{fig:supp_umap_ukbb_car}
\end{figure*}

\begin{figure*}
    \centering
    \begin{subfigure}{\textwidth}
    \includegraphics[width=\textwidth]{images/meanEuroMap.jpg}
    \end{subfigure}
    \caption{Figure~\ref{fig:umap_world_map}, zoomed in on Europe. Geographic distribution of UMAP coordinates. Using the country of birth of individuals in the UKBB, we colour countries by the closeness in 3D UMAP space of those born there. Broad patterns of similarity appear in East Asia, South Asia, North African and the Middle East, West Africa, and South America. Differences between neighbouring countries can reflect both ancient population structure and recent differences in migration history. Evidence of migrations related to colonialism are visible with, e.g., European ancestry in South Africa and South Asian ancestry in Kenya and Tanzania. Because of the large number of White British individuals born abroad, to avoid skewing the colour scale they were not included unless they were born in the UK, Europe, Australia, Canada, or the United States, where UKBB participants already tended to have European ancestry. }
    \label{fig:supp_umap_ukbb_eur}
\end{figure*}
